---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# survPen
## hazard and excess hazard modelling with multidimensional penalized splines
[![Travis-CI Build Status](https://travis-ci.org/fauvernierma/survPen.svg?branch=master)](https://travis-ci.org/fauvernierma/survPen)
[![Coverage Status](https://img.shields.io/codecov/c/github/fauvernierma/survPen/master.svg)](https://codecov.io/github/fauvernierma/survPen?branch=master)
[![DOI](https://zenodo.org/badge/181266005.svg)](https://zenodo.org/badge/latestdoi/181266005)

You can install this R package from GitHub:

```{r gh-installation, eval = FALSE}
install.packages("devtools")
library(devtools)
install_github("fauvernierma/survPen")
```

or directly from the CRAN repository:

```{r cran-installation, eval = FALSE}
install.packages("survPen")
```

and then use it by typing

```{r loading, eval = FALSE}
library(survPen)
```

## The functionalities of the *survPen* package are extensively detailed in the following vignette
Available from
[GitHub](https://htmlpreview.github.io/?https://github.com/fauvernierma/survPen/blob/master/inst/doc/survival_analysis_with_survPen.html)
or from [CRAN](https://cran.r-project.org/web/packages/survPen/vignettes/survival_analysis_with_survPen.html)

## Contributing

File an issue [here](https://github.com/fauvernierma/survPen/issues) if there is a feature, or a dataset, that you think is missing from the package, or better yet submit a pull request!

Please note that the `survPen` project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## Citing 

If using `survPen` please consider citing the package in the relevant work. Citation information can be generated in R using the following (after installing the package),

```{r}
citation("survPen")
```

You may also consider citing the method article:
Fauvernier, M., Roche, L., Uhry, Z., Tron, L., Bossard, N., Remontet, L. and the CENSUR Working Survival Group (2019). Multidimensional penalized hazard model with continuous covariates: applications for studying trends and social inequalities in cancer survival, in revision in the Journal of the Royal Statistical Society, series C.






